---
title: "Choose a limited sample of representative years"
author: "Jean Thorey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choose a limited sample of representative years}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = FALSE, 
  message = FALSE,
  warning = FALSE, 
  error = FALSE,
  fig.align = "center"
)
```


## Problem description

We want to find representative samples with block constraints. For example, we want to find a sample of 10 years amongst 50 years with a daily resolution. The daily resolution of the data must be preserved in the sample.

Monte Carlo method : this package implements a brute force approach to compute representativeness scores

TODO : add details about representativeness scores

## Example

Build data set

```{r, echo = TRUE}
library(lubridate)
library(data.table)
library(scenclimsample)

set.seed(2022)

# generate time series of daily data
nb_year <- 50
dt_series <- data.table(block_id = rep(seq(nb_year), each = 365),
                        horodate = rep(seq(ymd_h("2001-01-01 00"), ymd_h("2001-12-31 23"), by = "day"), nb_year),
                        value_1 = rnorm(365*nb_year),
                        value_2 = rnorm(365*nb_year))
```


Compute energy score matrices

```{r, echo = TRUE}
# build list of matrices A for energy score computation
l_mat_A <- purrr::map(c("value_1", "value_2"), function(ccol){
  build_euclid_dist_matrix(dt_series[,.SD,.SDcols = c("block_id", ccol)])
})
names(l_mat_A) <- c("value_1", "value_2")
```

Generate random samples

```{r, echo = TRUE}
# build random samples
Ntry <- 1e6
nb_year <- max(dt_series$block_id)
nb_year_by_sample <- 10

array_sample_random <- sapply(seq(Ntry), function(x) sample.int(nb_year, nb_year_by_sample, replace = FALSE))
colnames(array_sample_random) <- paste0("random_", seq(ncol(array_sample_random)))
```


Compute scores. Fast because specific implementation for uniform weights. TODO implement non-uniform weighting and Cpp implementation.

```{r, echo = TRUE}
# score computation
print(system.time({dt_score_avg <- get_dt_score_avg(array_sample = array_sample_random, dt_series = dt_series, col_vec = c("value_1", "value_2"))}))
print(system.time({dt_score_energy <- get_dt_score_energy(array_sample = array_sample_random, l_mat_A = l_mat_A)}))
print(system.time({dt_score_quantile <- get_dt_score_quantile(array_sample = array_sample_random[,1:1000], dt_series = dt_series, col_vec = c("value_1", "value_2"), upper_quantile_boundary = 0.99)}))
```

Check energy scores results

```{r, echo = TRUE}
# check vs. energy::edist
ws <- array(0, c(nb_year,1))
ws[array_sample_random[,1],1] <- 1/nrow(array_sample_random)
w0 <- array(1/nb_year, c(nb_year,1))

mat_A <- l_mat_A$value_1

# dt_score_energy
print(dt_score_energy[1,.(value_1)])

# dt_score_energy inner method
print(t(ws) %*% mat_A %*% w0 -  0.5*t(w0) %*% mat_A %*% w0 - 0.5*t(ws) %*% mat_A %*% ws)

# energy::edist
size_edist <- c(nrow(dt_series), nrow(dt_series[block_id %in% array_sample_random[,1]]))
edist_estim <- energy::edist(x = c(dt_series[,value_1],
                                   dt_series[block_id %in% array_sample_random[,1],value_1]),
                             sizes = size_edist) * sum(size_edist) / prod(size_edist) / 2
print(edist_estim)

```

