---
title: "Choose a limited sample of representative years"
author: "Jean Thorey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choose a limited sample of representative years}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.align = "center"
)
```

## Problem description

We want to extract representative samples of block time series (for example, select
a subset of years among a longer climate record) while preserving the intra-block
resolution (daily data inside each year). The package provides several scores to
evaluate how representative a candidate subset is compared to the full dataset.

The general workflow is:

1. Generate or import a block time series.
2. Precompute objects needed by the scores.
3. Draw candidate samples.
4. Compute scores and benchmark their durations.

## Example

### Build a dataset

```{r}
library(lubridate)
library(data.table)
library(scenclimsample)

set.seed(2022)

nb_year_total <- 30
nb_days <- 365

dt_series <- data.table(
  block_id = rep(seq_len(nb_year_total), each = nb_days),
  horodate = rep(seq(ymd_h("2001-01-01 00"), ymd_h("2001-12-31 23"), by = "day"), nb_year_total),
  value_1 = rnorm(nb_days * nb_year_total),
  value_2 = 0.5 * rnorm(nb_days * nb_year_total) + 0.2 * rep(seq_len(nb_year_total), each = nb_days)
)

dt_series
```

### Precompute objects for the scores

```{r}
col_vec <- c("value_1", "value_2")

# Euclidean distance matrices for energy scores
l_mat_A <- purrr::map(col_vec, function(ccol) {
  build_euclid_dist_matrix(dt_series[, .SD, .SDcols = c("block_id", ccol)])
})
names(l_mat_A) <- col_vec

# Block-wise CDFs for CDF scores
l_block_cdf <- build_l_block_cdf(dt_series, col_vec = col_vec)
```

### Generate random samples

```{r}
Ntry <- 2000
nb_year_by_sample <- 6

array_sample_random <- sapply(
  seq_len(Ntry),
  function(x) sample.int(nb_year_total, nb_year_by_sample, replace = FALSE)
)
colnames(array_sample_random) <- paste0("random_", seq_len(ncol(array_sample_random)))

array_sample_random[, 1:5]
```

### Compute all scores and benchmark durations

```{r}
timing_avg <- system.time({
  dt_score_avg <- get_dt_score_avg(
    array_sample = array_sample_random,
    dt_series = dt_series,
    col_vec = col_vec)
})

timing_energy <- system.time({
  dt_score_energy <- get_dt_score_energy(
    array_sample = array_sample_random,
    l_mat_A = l_mat_A)
})

timing_cdf <- system.time({
  dt_score_cdf <- get_dt_score_CDF(
    array_sample = array_sample_random,
    l_block_cdf = l_block_cdf,
    upper_quantile_boundary = 0.99)
})

timing_quantile <- system.time({
  dt_score_quantile <- get_dt_score_quantile(
    array_sample = array_sample_random,
    dt_series = dt_series,
    col_vec = col_vec,
    upper_quantile_boundary = 0.99)
})

benchmark <- data.table(
  score = c("avg", "energy", "cdf", "quantile"),
  elapsed_sec = c(
    as.numeric(timing_avg["elapsed"]),
    as.numeric(timing_energy["elapsed"]),
    as.numeric(timing_cdf["elapsed"]),
    as.numeric(timing_quantile["elapsed"])
  )
)
benchmark
```

### Inspect score outputs

```{r}

head(dt_score_avg)
head(dt_score_energy)
head(dt_score_quantile)
head(dt_score_cdf)

```

Each row corresponds to a candidate subset of years, and each column corresponds
to a score. Lower values indicate a subset whose distribution is closer to the
full dataset according to the corresponding metric.
