---
title: "Choose a limited sample of representative years"
author: "Jean Thorey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choose a limited sample of representative years}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.align = "center"
)
```

## Problem description

We want to extract representative samples of block time series (for example, select
a subset of years among a longer climate record) while preserving the intra-block
resolution (daily data inside each year). The package provides several scores to
evaluate how representative a candidate subset is compared to the full dataset.

The general workflow is:

1. Generate or import a block time series.
2. Precompute objects needed by the scores.
3. Draw candidate samples.
4. Compute scores.

## Score descriptions

Let \(j = 1, \ldots, N\) index the blocks (years). A candidate sample is encoded by
weights \(u \in \mathbb{R}^N\) (uniform on the selected years), and the full dataset
is represented by \(v = \left(\frac{1}{N}, \ldots, \frac{1}{N}\right)\). For a variable
\(X\), each block \(j\) defines a distribution \(F_j\).

### Mean score

Let \(m_j = \mathbb{E}[X \mid j]\) be the mean in block \(j\). The mean score compares
the sample mean to the full mean:
\[
S_{\text{avg}} = \left| u^\top m - v^\top m \right| .
\]

### Energy score

Let \(A\) be the matrix with entries \(A_{jk} = \mathbb{E}\left(\|X_j - X_k\|\right)\)
computed from block-wise Euclidean distances. The energy score is
\[
S_{\text{energy}} = u^\top A v - \frac{1}{2}u^\top A u - \frac{1}{2}v^\top A v .
\]

### CDF score

Define a grid \(x_1, \ldots, x_G\) (typically quantiles of the full dataset). The CDF
of the sample mixture is \(\widehat{F}_u(x_g) = \sum_j u_j F_j(x_g)\), and the CDF of
the full dataset is \(F_v(x_g) = \sum_j v_j F_j(x_g)\). The CDF score is
\[
S_{\text{cdf}} = \frac{1}{G}\sum_{g=1}^{G} \left| \widehat{F}_u(x_g) - F_v(x_g) \right| .
\]
A tail-focused version is obtained by restricting the grid to \(x_g\) above a
high quantile (e.g. \(x_g > q_{0.99}\)).

### Quantile score

Let \(\alpha_1, \ldots, \alpha_G\) be quantile levels and \(q_u(\alpha_g)\) (resp.
\(q_v(\alpha_g)\)) the quantiles of the sample mixture (resp. full dataset). The
quantile score is
\[
S_{\text{quantile}} = \frac{1}{G}\sum_{g=1}^{G} \left| q_u(\alpha_g) - q_v(\alpha_g) \right| .
\]

## Visual intuition for CDF and quantile scores

The two plots below compare the empirical CDFs and quantiles of a full dataset
and a candidate sample. The CDF score integrates the absolute difference between
the two CDFs on a grid, while the quantile score averages the absolute difference
between corresponding quantiles. The CDF and quantile scores differ in several aspects:

* **Grid discretization**
* **Tail scoring**
* **Computation time:** the CDF score is much faster because it relies on summation, whereas the quantile score requires sorting.





```{r}
library(ggplot2)
library(data.table)

set.seed(123)

full_values <- rnorm(1000, mean = 0, sd = 1)
sample_values <- rnorm(300, mean = 0.2, sd = 1.1)

grid_x <- as.numeric(quantile(full_values, probs = seq(0.01, 0.99, 0.01)))
cdf_full <- ecdf(full_values)(grid_x)
cdf_sample <- ecdf(sample_values)(grid_x)

dt_cdf <- data.table(
  x = rep(grid_x, 2),
  cdf = c(cdf_full, cdf_sample),
  dataset = rep(c("Full", "Sample"), each = length(grid_x))
)

p_cdf <- ggplot(dt_cdf, aes(x = x, y = cdf, color = dataset)) +
  geom_line(linewidth = 1) +
  labs(title = "Empirical CDF comparison", x = "X", y = "Probability") +
  theme_minimal()

alpha_grid <- seq(0.01, 0.99, 0.01)
q_full <- as.numeric(quantile(full_values, probs = alpha_grid))
q_sample <- as.numeric(quantile(sample_values, probs = alpha_grid))

dt_quantile <- data.table(
  alpha = rep(alpha_grid, 2),
  quantile = c(q_full, q_sample),
  dataset = rep(c("Full", "Sample"), each = length(alpha_grid))
)

p_quantile <- ggplot(dt_quantile, aes(x = alpha, y = quantile, color = dataset)) +
  geom_line(linewidth = 1) +
  labs(title = "Quantile comparison", x = "Probability", y = "X") +
  theme_minimal()

p_cdf
p_quantile
```

## Example

### Build a dataset

```{r}
library(lubridate)
library(data.table)
library(scenclimsample)

set.seed(2022)

nb_year_total <- 30
nb_days <- 365

dt_series <- data.table(
  block_id = rep(seq_len(nb_year_total), each = nb_days),
  horodate = rep(seq(ymd_h("2001-01-01 00"), ymd_h("2001-12-31 23"), by = "day"), nb_year_total),
  value_1 = rnorm(nb_days * nb_year_total),
  value_2 = 0.5 * rnorm(nb_days * nb_year_total) + 0.2 * rep(seq_len(nb_year_total), each = nb_days)
)

dt_series
```

### Precompute objects for the scores

```{r}
col_vec <- c("value_1", "value_2")

# Euclidean distance matrices for energy scores
l_mat_A <- purrr::map(col_vec, function(ccol) {
  build_euclid_dist_matrix(dt_series[, .SD, .SDcols = c("block_id", ccol)])
})
names(l_mat_A) <- col_vec

# Block-wise CDFs for CDF scores
l_block_cdf <- build_l_block_cdf(dt_series, col_vec = col_vec)
```

### Generate random samples

```{r}
Ntry <- 2000
nb_year_by_sample <- 6

array_sample_random <- sapply(
  seq_len(Ntry),
  function(x) sample.int(nb_year_total, nb_year_by_sample, replace = FALSE)
)
colnames(array_sample_random) <- paste0("random_", seq_len(ncol(array_sample_random)))

array_sample_random[, 1:5]
```

### Compute all scores and benchmark durations

```{r}
timing_avg <- system.time({
  dt_score_avg <- get_dt_score_avg(
    array_sample = array_sample_random,
    dt_series = dt_series,
    col_vec = col_vec)
})

timing_energy <- system.time({
  dt_score_energy <- get_dt_score_energy(
    array_sample = array_sample_random,
    l_mat_A = l_mat_A)
})

timing_cdf <- system.time({
  dt_score_cdf <- get_dt_score_CDF(
    array_sample = array_sample_random,
    l_block_cdf = l_block_cdf,
    upper_quantile_boundary = 0.99)
})

timing_quantile <- system.time({
  dt_score_quantile <- get_dt_score_quantile(
    array_sample = array_sample_random,
    dt_series = dt_series,
    col_vec = col_vec,
    upper_quantile_boundary = 0.99)
})

benchmark <- data.table(
  score = c("avg", "energy", "cdf", "quantile"),
  elapsed_sec = c(
    as.numeric(timing_avg["elapsed"]),
    as.numeric(timing_energy["elapsed"]),
    as.numeric(timing_cdf["elapsed"]),
    as.numeric(timing_quantile["elapsed"])
  )
)
benchmark
```

### Inspect score outputs

```{r}

head(dt_score_avg)
head(dt_score_energy)
head(dt_score_quantile)
head(dt_score_cdf)

```

Each row corresponds to a candidate subset of years, and each column corresponds
to a score. Lower values indicate a subset whose distribution is closer to the
full dataset according to the corresponding metric.
